---
title: "Elbow flexion_R_frontal"
author: "TZ"
date: "2026-04-18"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
set.seed(42)
```


# 0. Preparation
## loadPackages, message=FALSE, echo=FALSE
```{r}
library(lme4)
library(mgcv)
library(itsadug)
library(gratia)
library(ggplot2)
library(ggeffects)
library(readxl)
library(wesanderson)
library(dplyr)
library(tidyr)
library(emmeans)
library(nlme)
library(car)
library(patchwork)
```

# GAM-CPU distribution - use half of the available cores
```{r}
n.cores <- max(1, parallel::detectCores() %/% 2, na.rm = T)
```


# Read data
```{r}
df <- read_excel("D:/next.Hessenbox/Human-technology interaction/Motion tracking accuracy comparison/Data_Zenodo/Elbow_flexion_R.xlsx")
head (df)
```


# 1. Calculate MAE and RMSE
```{r}
mae  <- function(x, y, na.rm = TRUE) mean(abs(x - y), na.rm = na.rm)
rmse <- function(x, y, na.rm = TRUE) sqrt(mean((x - y)^2, na.rm = na.rm))

long_err <- df %>%
  pivot_longer(c(Kinect, Captiv), names_to = "device", values_to = "angle") %>%
  mutate(
    abs_err = abs(angle - Angle_scale),
    sq_err  = (angle - Angle_scale)^2
  )

summarizer <- function(.data, ...) {
  grp <- quos(...)
  .data %>%
    { if (length(grp) > 0) group_by(., !!!grp, device) else group_by(., device) } %>%
    summarise(
      n    = n(),
      MAE  = mean(abs_err, na.rm = TRUE),
      RMSE = sqrt(mean(sq_err, na.rm = TRUE)),
      .groups = "drop_last"
    ) %>%
    pivot_wider(
      names_from = device,
      values_from = c(MAE, RMSE, n),
      names_sep = "_"
    ) %>%
    mutate(
      dMAE  = MAE_Captiv  - MAE_Kinect,
      dRMSE = RMSE_Captiv - RMSE_Kinect
    ) %>%
    { if (length(grp) > 0) arrange(., !!!grp) else . }
}

overall_tbl <- summarizer(long_err) %>% mutate(level = "overall", .before = 1)
by_part_tbl <- summarizer(long_err, Participant) %>% mutate(level = "by_participant", .before = 1)
by_pt_tbl   <- summarizer(long_err, Participant, Trial) %>% mutate(level = "by_participant_trial", .before = 1)

comparison_all_levels <- bind_rows(overall_tbl, by_part_tbl, by_pt_tbl)
comparison_all_levels
```



# 2. GAM unfactored approach 
## 2.1 Reorganize dataset
```{r}
df <- df %>%
  mutate(Kinect_error = Kinect - Angle_scale,
         Captiv_error = Captiv - Angle_scale)

# Reshape
df_long <- df %>%
  pivot_longer(cols = c(Kinect_error, Captiv_error),
               names_to = "Device", 
               values_to = "Error")

head(df_long)
```


## 2.2 Factorize categorical variables
```{r}
df_long$Participant <- factor(df_long$Participant, ordered=F)
df_long$Trial <- factor(df_long$Trial, ordered=F)
df_long$Device <- factor(df_long$Device, levels = c("Kinect_error", "Captiv_error"), ordered=F)
str(df_long)
```


## 2.3a Create Trial ID as a new variable that integrates trial and participant
```{r}
df_long <- df_long %>%
  mutate(TrialID = interaction(Participant, Trial, drop = TRUE))
df_long$TrialID <- factor(df_long$TrialID, ordered=F)
str(df_long)
```

## 2.3b Create SeriesID and time index for autocorrelation diagnostics
```{r}
df_long <- df_long %>%
  mutate(
    SeriesID = interaction(Participant, Trial, Device, drop = TRUE)
  ) %>%
  arrange(SeriesID, Frame) %>%
  group_by(SeriesID) %>%
  mutate(
    t = row_number()
  ) %>%
  ungroup()

str(df_long)
```


## 2.3c Helper functions for diagnostics
```{r}
acf1 <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 3) return(NA_real_)
  acf(x, plot = FALSE, na.action = na.pass)$acf[2]
}

plot_qq <- function(diag_df, title = "QQ plot") {
  ggplot(diag_df, aes(sample = resid)) +
    stat_qq(alpha = 0.4) +
    stat_qq_line() +
    theme_bw() +
    labs(title = title, x = "Theoretical quantiles", y = "Sample quantiles")
}

plot_resid_fitted <- function(diag_df, title = "Residuals vs fitted") {
  ggplot(diag_df, aes(x = fitted, y = resid)) +
    geom_point(alpha = 0.3) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_smooth(method = "loess", se = TRUE) +
    theme_bw() +
    labs(title = title, x = "Fitted values", y = "Residuals")
}

plot_scale_location <- function(diag_df, title = "Scale-location plot") {
  ggplot(diag_df, aes(x = fitted, y = sqrt(abs(resid)))) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "loess", se = TRUE) +
    theme_bw() +
    labs(title = title, x = "Fitted values", y = "sqrt(|Residuals|)")
}

plot_lag1_acf_hist <- function(diag_df, title = "Lag-1 residual ACF") {
  rho_df <- diag_df %>%
    group_by(SeriesID) %>%
    summarise(rho1 = acf1(resid), .groups = "drop")

  p <- ggplot(rho_df, aes(x = rho1)) +
    geom_histogram(binwidth = 0.05) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    theme_bw() +
    labs(title = title, x = "Lag-1 residual ACF", y = "Count")

  list(plot = p, rho_df = rho_df)
}

# VIF only for parametric part
get_parametric_vif <- function(model, data) {
  tt <- delete.response(terms(model$pterms))
  X <- model.matrix(tt, data = data)

  # remove intercept
  if ("(Intercept)" %in% colnames(X)) {
    X <- X[, colnames(X) != "(Intercept)", drop = FALSE]
  }

  if (ncol(X) < 2) {
    return(data.frame(term = colnames(X), VIF = NA_real_))
  }

  vif_vals <- car::vif(lm(X[, 1] ~ X[, -1, drop = FALSE]))
  data.frame(term = names(vif_vals), VIF = as.numeric(vif_vals))
}

extract_concurvity <- function(model, model_name) {
  cc <- concurvity(model, full = TRUE)
  out <- as.data.frame(cc)
  out$term <- rownames(out)
  rownames(out) <- NULL
  out$model <- model_name
  out %>%
    select(model, term, worst, observed, estimate)
}
```


## 2.4 GAM model using unordered factor approach
```{r}
m.tid <- gam(Error 
         ~ 1                              # Parametric intercept
         + Device                         # Parametric main effect of Group
         + s(Angle_scale, by=Device)      # Smooth for angle scale (separately for each Device)
         + s(Participant, bs="re")        # By-participant random intercepts
         + s(TrialID, bs="re"),           # By-trial random intercepts
         data = df_long,
         method = "REML",
)
```


## 2.5 Prediction using unordered factor approach
```{r}
ggemmeans(m.tid, c("Angle_scale [all]", "Device")) |>
  plot() +
  scale_color_manual(values = c("Kinect_error" = "red", "Captiv_error" = "blue")) +
  scale_fill_manual(values = c("Kinect_error" = "red", "Captiv_error" = "blue"))
```


## 2.6 Model summary
```{r}
## ----gam-both-sum-par---------------------------
summary(m.tid)$p.table


## ----gam-both-sum-smo---------------------------
summary(m.tid)$s.table
```


## 2.7 k check
```{r}
k.check(m.tid, n.rep = 10^3)
```


## 2.8 Residual diagnostics for m.tid
```{r}
diag_tid <- df_long %>%
  mutate(
    fitted = fitted(m.tid),
    resid  = residuals(m.tid, type = "deviance")
  )

p_tid_qq <- plot_qq(diag_tid, "model(1): QQ plot")
p_tid_rf <- plot_resid_fitted(diag_tid, "model(1): Residuals vs fitted")
p_tid_sl <- plot_scale_location(diag_tid, "model(1): Scale-location")
acf_tid  <- plot_lag1_acf_hist(diag_tid, "model(1): Lag-1 residual ACF")

p_all <- p_tid_qq | p_tid_rf | p_tid_sl | acf_tid$plot
ggsave(
  filename = "Elbow_flexion_R_model1_diagnostics.svg",
  plot = p_all,
  width = 20,
  height = 5,
  device = svglite::svglite
)

median(acf_tid$rho_df$rho1, na.rm = TRUE)
```


## 2.9 Concurvity and VIF for m.tid
```{r}
# Concurvity
extract_concurvity <- function(model, model_name) {
  cc <- concurvity(model, full = TRUE)
  cc_df <- as.data.frame(t(cc))
  cc_df$term <- rownames(cc_df)
  rownames(cc_df) <- NULL
  cc_df$model <- model_name

  cc_df %>%
    select(model, term, worst, observed, estimate)
}

concurvity_tid <- extract_concurvity(m.tid, "m.tid")
concurvity_tid

# VIF for parametric part only
vif_tid <- get_parametric_vif(m.tid, df_long)
vif_tid
```



# 3. GAM model with factor smooth
## 3.1 GAM model with factor smooths
```{r}
m.fs <- gam(Error 
         ~ 1                                       # Parametric intercept
         + Device                                  # Parametric main effect of Group
         + s(Angle_scale, by=Device)               # Smooth for angle scale (separately for each Device)
         + s(Angle_scale, Participant, bs="fs")    # Factor smooths
         + s(TrialID, bs="re"),                    # By-trialID random intercepts
         data = df_long,
         method = "REML",
)
```


```{r}
summary(m.fs)
```


## 3.2 Model prediction
```{r}
ggemmeans(m.fs, c("Angle_scale [all]", "Device")) |>
  plot() +
  scale_color_manual(values = c("Kinect_error" = "red", "Captiv_error" = "blue")) +
  scale_fill_manual(values = c("Kinect_error" = "red", "Captiv_error" = "blue"))
```


## 3.3 Model summary of factor smooths approach
```{r}
## ----gam-both-sum-par---------------------------
summary(m.fs)$p.table


## ----gam-both-sum-smo---------------------------
summary(m.fs)$s.table
```


## 3.4 k check
```{r}
k.check(m.fs, n.rep = 10^3)
```
### Interpretation: p value is not significant, indicating that k is sufficient.


## 3.5 Compare factor smooth model and previous model
```{r}
compareML(m.tid, m.fs)
```
### Interpretation: AIC for m.fs is lower, indicating that model with factor smooth is a better model.


## 3.6 Residual diagnostics for m.fs
```{r}
diag_fs <- df_long %>%
  mutate(
    fitted = fitted(m.fs),
    resid  = residuals(m.fs, type = "deviance")
  )

p_fs_qq <- plot_qq(diag_fs, "model(2): QQ plot")
p_fs_rf <- plot_resid_fitted(diag_fs, "model(2): Residuals vs fitted")
p_fs_sl <- plot_scale_location(diag_fs, "model(2): Scale-location")
acf_fs  <- plot_lag1_acf_hist(diag_fs, "model(2): Lag-1 residual ACF")

p_all <- p_fs_qq | p_fs_rf | p_fs_sl | acf_fs$plot
ggsave(
  filename = "Elbow_flexion_R_model2_diagnostics.svg",
  plot = p_all,
  width = 20,
  height = 5,
  device = svglite::svglite
)

median(acf_fs$rho_df$rho1, na.rm = TRUE)
```


## 3.7 Concurvity and VIF for m.fs
```{r}
# Concurvity

concurvity_fs <- extract_concurvity(m.fs, "m.fs")
concurvity_fs

# VIF for parametric part only
vif_fs <- get_parametric_vif(m.fs, df_long)
vif_fs
```



# 4. Consider auto-correlation in the GAM model
## 4.1 Define nested trial and the AR-start flag
```{r}
df_long %>%
  select(Participant, Trial, Device, TrialID, SeriesID, Frame, t) %>%
  arrange(SeriesID, Frame) %>%
  head(20)
```


## 4.2 Visualize the residuals to see whether thers is auto-correlation
```{r}
# residuals from factor-smooth model without AR(1)
df_long$resid_fs <- residuals(m.fs, type = "deviance")

rho_df <- df_long %>%
  group_by(SeriesID) %>%
  summarise(rho1 = acf1(resid_fs), .groups = "drop")

ggplot(rho_df, aes(x = rho1)) +
  geom_histogram(binwidth = 0.05) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw() +
  labs(title = "m.fs: Lag-1 ACF of residuals", x = "rho1", y = "Count")

median(rho_df$rho1, na.rm = TRUE)
```
### Interpretation: the histogram is not centered around 0, most values are larger than 0.5, thus auto-correlation is likely existed.


## 4.3 Fit AR(1) into GAM
```{r}
m.gamm_AR1 <- gamm(
  Error ~ 
    Device 
  + s(Angle_scale, by = Device) 
  + s(Angle_scale, Participant, bs = "fs"),
  random = list(TrialID = ~ 1),
  correlation = corAR1(form = ~ t | SeriesID),   # AR(1) within Trial×Device
  data   = df_long,
  method = "REML"
)
```


## 4.4 Model prediction
```{r}
pdat <- ggpredict(
  m.gamm_AR1$gam,
  terms   = c("Angle_scale [all]", "Device"),
  exclude = c("s(Angle_scale,Participant)", "s(TrialID)")  # key line
)

p <- plot(pdat) +
  labs(
    x = "Angle scale benchmark value",                 
    y = "Predicted measurement difference",          
    title = "Right elbow flexion" 
  ) +
  scale_color_manual(
    name   = "Device",   
    values = c(Kinect_error = "red", Captiv_error = "blue"),
    labels = c("Azure Kinect", "Captiv") 
  ) +
  scale_fill_manual(
    name   = "Device",
    values = c(Kinect_error = "red", Captiv_error = "blue"),
    labels = c("Azure Kinect", "Captiv")
  ) +
  scale_x_continuous(breaks = seq(0, 150, by = 10), limits = c(0,150)) + 
  scale_y_continuous(breaks = seq(-60, 40, by = 10), limits = c(-60,40)) + 
  coord_fixed(ratio = 1)

p <- p + theme(
  plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "cm") 
)

print(p)

ggsave(filename = "Elbow_flexion_R_model3_prediction.svg", plot = p, width = 8, height = 6, device = svglite::svglite)

```


## 4.5 Model summary
```{r}
## ----gam-both-sum-par---------------------------
summary(m.gamm_AR1$gam)$p.table


## ----gam-both-sum-smo---------------------------
summary(m.gamm_AR1$gam)$s.table

## AR(1)
summary(m.gamm_AR1$lme)  
```


## 4.6 Residual diagnostics for m.gamm_AR1
```{r}
diag_ar1 <- df_long %>%
  mutate(
    fitted = fitted(m.gamm_AR1$lme),
    resid  = resid(m.gamm_AR1$lme, type = "normalized")
  )

p_ar1_qq <- plot_qq(diag_ar1, "model(3): QQ plot")
p_ar1_rf <- plot_resid_fitted(diag_ar1, "model(3): Residuals vs fitted")
p_ar1_sl <- plot_scale_location(diag_ar1, "model(3): Scale-location")
acf_ar1  <- plot_lag1_acf_hist(diag_ar1, "model(3): Lag-1 residual ACF")

p_all <- p_ar1_qq | p_ar1_rf | p_ar1_sl | acf_ar1$plot
ggsave(
  filename = "Elbow_flexion_R_model3_diagnostics.svg",
  plot = p_all,
  width = 20,
  height = 5,
  device = svglite::svglite
)

median(acf_ar1$rho_df$rho1, na.rm = TRUE)
```


## 4.7 Model comparison
```{r}
m.gamm_noAR <- gamm(
  Error ~ Device + s(Angle_scale, by = Device) + s(Angle_scale, Participant, bs = "fs"),
  random = list(TrialID = ~ 1),
  data   = df_long, method = "REML"
)

m.gamm_AR1 <- gamm(
  Error ~ Device + s(Angle_scale, by = Device) + s(Angle_scale, Participant, bs = "fs"),
  random = list(TrialID = ~ 1),
  correlation = corAR1(form = ~ t | SeriesID),
  data   = df_long, method = "REML"
)

## Model comparison
anova(m.gamm_noAR$lme, m.gamm_AR1$lme)  
AIC(m.gamm_noAR$lme, m.gamm_AR1$lme)
```
### Interpretation: factor smooths with AR(1) term is the best model so far.


## 4.8 Diagnostic summary table
```{r}
diag_summary <- bind_rows(
  acf_tid$rho_df %>%
    summarise(Model = "m.tid",
              median_rho1 = median(rho1, na.rm = TRUE),
              mean_rho1   = mean(rho1, na.rm = TRUE)),
  acf_fs$rho_df %>%
    summarise(Model = "m.fs",
              median_rho1 = median(rho1, na.rm = TRUE),
              mean_rho1   = mean(rho1, na.rm = TRUE)),
  acf_ar1$rho_df %>%
    summarise(Model = "m.gamm_AR1",
              median_rho1 = median(rho1, na.rm = TRUE),
              mean_rho1   = mean(rho1, na.rm = TRUE))
)

diag_summary
```


# 5. Bland-Altman analysis
## 5.1 Prepare raw BA dataset
```{r}
ba_raw <- bind_rows(
  df %>%
    transmute(
      Participant,
      Trial,
      Device = "Azure Kinect",
      Mean = (Kinect + Angle_scale) / 2,
      Diff = Kinect - Angle_scale
    ),
  df %>%
    transmute(
      Participant,
      Trial,
      Device = "Captiv",
      Mean = (Captiv + Angle_scale) / 2,
      Diff = Captiv - Angle_scale
    )
) %>%
  filter(!is.na(Mean), !is.na(Diff))

head(ba_raw)
```


## 5.2 Compute Bland-Altman bias and limits of agreement
```{r}
ba_limits <- ba_raw %>%
  group_by(Device) %>%
  summarise(
    n = n(),
    Bias = mean(Diff, na.rm = TRUE),
    SD_diff = sd(Diff, na.rm = TRUE),
    LoA_lower = Bias - 1.96 * SD_diff,
    LoA_upper = Bias + 1.96 * SD_diff,
    .groups = "drop"
  )

ba_limits
```


## 5.3 Bland-Altman plot
```{r}
p_ba <- ggplot(ba_raw, aes(x = Mean, y = Diff)) +
  geom_point(alpha = 0.20, size = 0.8) +
  facet_wrap(~ Device, ncol = 2) +
  geom_hline(
    data = ba_limits,
    aes(yintercept = Bias),
    linetype = "solid",
    inherit.aes = FALSE
  ) +
  geom_hline(
    data = ba_limits,
    aes(yintercept = LoA_lower),
    linetype = "dashed",
    inherit.aes = FALSE
  ) +
  geom_hline(
    data = ba_limits,
    aes(yintercept = LoA_upper),
    linetype = "dashed",
    inherit.aes = FALSE
  ) +
  theme_bw() +
  labs(
    title = "Bland-Altman plots",
    x = "Mean of device and angle scale benchmark [°]",
    y = "Difference (device - angle scale benchmark) [°]"
  )

print(p_ba)
ggsave("Elbow_flexion_R_Bland_Altman_raw.svg", p_ba, width = 13, height = 5)
```
